Mainstreaming Race Science

### 2023-09-26 paper/paper.Rmd
updating paper and refs-manual
I've updated the "Race science, behavior genetics, and psychology" section to include more text relevant to psychology's lack of forceful repudiation of race science relative to other disciplines, and updated refs-manual to include the new references for that text.

| | | |
|:-----|-----:|----:|
| Thomas J. Bouchard, Jr. (1937-) | psychology | 184 |
| Brunetto Chiarelli (?-?) | anthropology | 91 |
| Hans Eysenck (1916-1997) | psychology | 661 |
| Robert A. Gordon (1932-) | sociology | 0|
| Linda Gottfredson (1947-) | psychology | 71 |
| Garrett Hardin (1915-2003) | ecology | 75 |
| Joseph M. Horn (1940-) | psychology | 68 |
| Lloyd Humphreys (1913-2003) | psychology | 0 |
| Arthur Jensen (1923-2003) | psychology | 235 |
| Michael Levin (1943-) | philosophy | 96 |
| Richard Lynn (1930-) | psychology | 288 |
| R. Travis Osborne (1913-2013) | psychology | 59 |
| J. Phillippe Rushton (1943-2012) | psychology | 277 |
| Audrey M. Shuey (1900-1977) | psychology | 10 |
| Philip A. Vernon (1950-) | psychology | 227 |
| Daniel Vining, Jr. (1944-) | demography | 33 |

@@

However, the topic model analysis suggests that by the 1990s, behavior genetics may have distanced itself somewhat from race science, albeit without directly repudiating it. This interpretation is further supported by two disciplinary histories of behavior genetics, one published in McGue in 2010 \cite{McGueEndBehavioralGenetics2010} and the other by Plomin in 2023 \cite{PlominCelebratingCenturyResearch2023}. In addition, the 1995 BGA presidential address by Glayde Whitney --- in which he criticized the "Marixist-Lysenkoist denial of genetics" and proposed that differences in murder rates between countries and cities was caused by racial genetic differences in intelligence, empathy, aggression, and impulsivity --- was published in MQ rather than BG. As a discipline, behavior genetics may have been prominent in forming an echo chamber for race science in the past, but appears to have been less receptive to such ideas in more recent history.

Instead, our results suggest that psychology has been the dominant discipline in which the race science echo chamber has been maintained. This may be the result of racial essentialization being maintained more easily in a discipline like psychology than in other disciplines that contribute to behavior genetics, like genetics and anthropology.  The typical practitioners in psychology do not have the degree of formal education about, and therefore knowledge of, relevant underlying genetic and anthropological research that has led many professional organizations in genetics and anthropology to formally reject race as a biologically meaningful concept \cite{anthropology1996aapa, american2018ashg, board1998aaa, FuentesAAPAStatementRace2019, nhgri2023, rotimi2022ashg, wynshawboris2020ashg}, undermining the core assumption of race science. As described by Winston, however, "[s]ince the 1990s, scientific racism has evolved in ways that sidestep or blunt the charge of essentialist categorization" \cite[p. 427]{WinstonWhyMainstreamResearch2020}, such as shifting the rhetoric away from race to concepts such as "national IQ" that are deployed to preserve an essentialist categorization of different human populations. Professional psychological organizations routinely publish statements rejecting racism or acknowledging psychology's complicity in systemic racism \cite{american2021apology, aps2021statement}, but stop short of rejecting race as a meaningful biological categorization of people, allowing for race science research to be smuggled into psychology and giving the appearance of continued scientific legitimacy.

>>>>

| | | |
|:-----|-----:|----:|
| Thomas J. Bouchard, Jr. (1937-) | psychology | 184 |
| Brunetto Chiarelli (?-?) | anthropology | 91 |
| Hans Eysenck (1916-1997) | psychology | 661 |
| Robert A. Gordon (1932-) | sociology | 0|
| Linda Gottfredson (1947-) | psychology | 71 |
| Garrett Hardin (1915-2003) | ecology | 75 |
| Joseph M. Horn (1940-) | psychology | 68 |
| Lloyd Humphreys (1913-2003) | psychology | 0 |
| Arthur Jensen (1923-2003) | psychology | 235 |
| Michael Levin (1943-) | philosophy | 96 |
| Richard Lynn (1930-2023) | psychology | 288 |
| R. Travis Osborne (1913-2013) | psychology | 59 |
| J. Phillippe Rushton (1943-2012) | psychology | 277 |
| Audrey M. Shuey (1900-1977) | psychology | 10 |
| Philip A. Vernon (1950-) | psychology | 227 |
| Daniel Vining, Jr. (1944-) | demography | 33 |

@@

However, the topic model analysis suggests that by the 1990s, behavior genetics may have distanced itself somewhat from race science, albeit without directly repudiating it. This interpretation is further supported by two disciplinary histories of behavior genetics, one published in McGue in 2010 \cite{McGueEndBehavioralGenetics2010} and the other by Plomin in 2023 \cite{PlominCelebratingCenturyResearch2023}. In addition, the 1995 BGA presidential address by Glayde Whitney --- in which he criticized the "Marixist-Lysenkoist denial of genetics" and proposed that differences in murder rates between countries and cities was caused by racial genetic differences in intelligence, empathy, aggression, and impulsivity --- was published in MQ rather than BG. As a subdiscipline of psychology, behavior genetics may have been prominent in promoting race science in the past, but appears to have been less receptive to such ideas in more recent history.

Instead, our results suggest that journals focusing subdisciplines within psychology beyond behavior genetics have been the environments in which the mainstreaming of race science has taken place. This may be the result of racial essentialization being maintained more easily in a discipline like psychology than in other disciplines that contribute to behavior genetics, like genetics and anthropology. The typical practitioners in psychology do not have the degree of formal education about, and therefore knowledge of, relevant underlying genetic and anthropological research that has led many professional organizations in genetics and anthropology to formally and routinely reject race as a biologically meaningful concept over the past several decades \cite{anthropology1996aapa, american2018ashg, board1998aaa, committee2020asa, FuentesAAPAStatementRace2019, nhgri2023, rotimi2022ashg, wynshawboris2020ashg}, undermining the core assumption of race science.

As described by Winston, however, "[s]ince the 1990s, scientific racism has evolved in ways that sidestep or blunt the charge of essentialist categorization" \cite[p. 427]{WinstonWhyMainstreamResearch2020}, such as shifting the rhetoric away from race to concepts such as "national IQ" that are deployed to preserve an essentialist categorization of different human populations. As such, another way to view the role that psychology has played in maintaining the legitimacy of race science is by attending to top-down influences within the profession. Professional psychological organizations routinely publish statements rejecting racism or acknowledging psychology's historical complicity in systemic racism, but do not uniformly reject race as a biological concept. The American Psychological Association (APA), whose members are predominantly clinical psychologists, and the Federation of Associations in Behavioral & Brain Sciences (FABBS), an organization whose purpose is to provide policy recommendations, have openly rejected race as biologically meaningful \cite{american2021apology, baron2022fabbs} whereas research-focused psychological organizations like the Association for Psychological Science (APS) and Psychonomic Society (PS) stop short of rejecting race as a biologically meaningful when they denounce racism\cite{aps2021statement, board2020psychonomic}. The lack of a uniform professional rejection of the core claim of race science allows for race science research to be smuggled into mainstream psychological research. The hesitancy of professional organizations like APS or PS are potentially due to the prominence with which several notable race scientists have achieved in scientific gate keeping positions, such as on journal editorial boards. Additionally, psychology as a discipline routinely upholds several mythological narratives concerning race science and the history of psychology, such as allegations of a taboo about race and intelligence research\cite{jackson2021mythical}, claims that scientists engaging in or calling for any kind of anti-racism in the field is an ideological corruption of dispassionate and value-neutral science\cite{roberts_2022}, or arguments that holding race science to the kinds of evidentiary standards that other psychological research is held to is somehow a violation of academic freedom\cite{herbert2023academic}.
